A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-20 main.tex
Sections 2.1 and 2.2

This paper presents the first full mechanical formalization of the metatheory for
higher-ranked polymorphic type inference. The system that we formalize
is the bidirectional type system by Dunfield and
Krishnaswami (DK). The DK type system has two variants (a declarative
and an algorithmic one) that have been \emph{manually} proven
\emph{sound}, \emph{complete} and \emph{decidable}. We present 
a mechanical formalization in the Abella theorem prover 
of DK's declarative type system with a novel algorithmic system. We have
several reasons to use a new algorithm. Firstly, our new algorithm
employs \emph{worklist judgements}, which allow precise capture of the
scope of variables and simplify the formalization of scoping in a
theorem prover. Secondly, while DK's original formalization comes with very
well-written manual proofs, there are several details missing,
which complicate the task of writing a formal proof. 
Moreover, some proofs/lemmas are wrong and it is
not clear how to replace these by other valid ones. 
Despite the use of a different algorithm we prove the
same results as DK, although with significantly different proofs and
proof techniques. Since such type inference algorithms are quite
subtle and have a complex metatheory, we believe that mechanical
formalizations are an important advance in type-inference research.
\end{abstract}

@@

%% Keywords
%% comma separated list
\keywords{type inference, higher-rank polymorphism, mechanisation}  %% \keywords are mandatory in final camera-ready submission

>>>>

This paper presents the first full mechanical formalization of the metatheory for
higher-ranked polymorphic type inference. The system that we formalize
is the bidirectional type system by Dunfield and
Krishnaswami (DK). The DK type system has two variants (a declarative
and an algorithmic one) that have been \emph{manually} proven
\emph{sound}, \emph{complete} and \emph{decidable}. We present 
a mechanical formalization in the Abella theorem prover 
of DK's declarative type system with a novel algorithmic system. We have
several reasons to use a new algorithm. Firstly, our new algorithm
employs \emph{worklist judgements}, which allow precise capture of the
scope of variables and simplify the formalization of scoping in a
theorem prover. Secondly, while DK's original formalization comes with very
well-written manual proofs, there are several details missing,
which complicate the task of writing a formal proof. 
Moreover, some proofs/lemmas are wrong and it is
not clear how to replace these by other valid ones. 
Despite the use of a different algorithm we prove the
same results as DK, although with significantly different proofs and
proof techniques. Since such type inference algorithms are quite
subtle and have a complex metatheory, mechanical
formalizations are an important advance in type-inference research.
\end{abstract}

@@

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

### 2019-02-20 sections/Overview.tex
Sections 2.1 and 2.2

\bruno{I think some text/ideas-for-text can be borrowed from ITP.}

@@

The syntax of this declarative system is shown in Figure~\ref{fig:decl:syntax}.
A declarative type $A$ is either the unit type $1$, a type variable $a$,
a universal quantification $\forall a. A$ or a function type $A \to B$.
Nested universal quantifiers are allowed for types,
but monotypes $\tau$ do not have any universal quantifier.
Terms include a unit term $()$, variables $x$, lambda-functions $\lam e$,
applications $e_1~e_2$ and annotations $(e:A)$.
Contexts $\Psi$ are sequences of type variable declarations and
term variables with its type declared $x:A$.

@@

\begin{figure}[t]
\begin{tabular}{rl}
    \framebox{$\Psi \vdash A \Lto B$} & $e$ checks against input type $A$.\\[0.5mm]
    \framebox{$\Psi \vdash A \To B$} & $e$ synthesizes output type $A$.\\[0.5mm]
    \framebox{$\Psi \vdash \appInf{A}{e}{C}$} & Applying a function of type $A$ to $e$ synthesizes type $C$.
\end{tabular}
\begin{gather*}
\inferrule*[right=$\mathtt{DeclVar}$]
    {(x:A)\in\Psi}{\Psi\vdash x\To A}
\qquad
\inferrule*[right=$\mathtt{DeclSub}$]
%e \neq \lam e' \quad B \neq \all B' \quad 
    {\Psi\vdash e\To A \quad \Psi\vdash A\le B}
    {\Psi \vdash e\Lto B}
\\
\inferrule*[right=$\mathtt{DeclAnno}$]
    {\Psi \vdash A \quad \Psi\vdash e\Lto A}
    {\Psi\vdash (e:A)\To A}
\qquad
\inferrule*[right=$\mathtt{Decl1I}$]
    {~}{\Psi\vdash () \Lto 1}
\qquad
\inferrule*[right=$\mathtt{Decl1I\To}$]
    {~}{\Psi\vdash () \To 1}
\\
\inferrule*[right=$\mathtt{Decl\forall I}$]
    {\Psi,a \vdash e \Lto A}
    {\Psi\vdash e\Lto \all A}
\qquad
\inferrule*[right=$\mathtt{Decl\forall App}$]
    {\Psi \vdash \tau \quad \Psi\vdash \appInf{[\tau/a]A}{e}{C} }
    {\Psi\vdash \appInf{\all A}{e}{C}}
\\
\inferrule*[right=$\mathtt{Decl\to I}$]
    {\Psi,x:A \vdash e\Lto B}
    {\Psi\vdash \lam e \Lto A \to B}
\qquad
\inferrule*[right=$\mathtt{Decl\to I\To}$]
    {\Psi\vdash \sigma\to\tau \quad \Psi,x:\sigma \vdash e\Lto \tau}
    {\Psi\vdash \lam e \To \sigma\to\tau}
\\
\inferrule*[right=$\mathtt{Decl\to E}$]
    {\Psi\vdash e_1\To A \quad \Psi\vdash \appInf{A}{e_2}{C}}
    {\Psi\vdash e_1~e_2 \To C}
\qquad
\inferrule*[right=$\mathtt{Decl\to App}$]
    {\Psi\vdash e \Lto A}
    {\Psi\vdash \appInf{A \to C}{e}{C}}
\end{gather*}
\caption{Declarative Typing}\label{fig:decl:typing}
\end{figure}

Figures~\ref{fig:decl:sub} and \ref{fig:decl:typing},
give the subtyping and typing relations for the declarative DK system.
The DK subtyping relation was adopted from \citet{odersky1996putting}.

@@

The most interesting rule is $\mathtt{{\le}\forall L}$.
If some instantiation of $\all A$, $[\tau/a]A$, is a subtype of $B$,
then $\all A \le B$.
The monotype $\tau$ we used to instantiate $a$ is \emph{guessed} in this
declarative rule, but the algorithmic system does not guess and defers the
instantiation until it can determine the monotype deterministically.
The fact that $\tau$ is a monotype rules out the possibility of
polymorphic (or impredicative) instantiation.\tom{say why this restriction is imposed}
Peyton Jones et al.~\cite{} also impose the restriction of
predicative instantiation in their type system.
Both systems are adopt by several practical programming languages.

@@

Some of the rules handles simple cases with common approaches.
Rule $\mathtt{DeclVar}$ looks term variables up in the context.
Rules $\mathtt{Decl1I}$ and $\mathtt{Decl1I{\To}}$ checks and synthesizes the unit type, respectively.
Rule $\mathtt{DeclAnno}$ synthesizes the type of the annotated expression $(e:A)$
to its annotated type $A$, and requires $e$ to be checked agains $A$.
Checking an expression $e$ against a polymorphic type $\all A$ under the context $\Psi$ succeeds
if $e$ checks against A under an extended context $(\Psi, a)$, where $a$ is picked as a free variable.

The subsumption rule $\mathtt{DeclSub}$ depends on the subtyping relation,
and changes mode from checking to synthesis: if $e$ synthesizes $A$ and $A\le B$
($A$ is more polymorphic than $B$), then $e$ can be checked against $B$.
If a checking problem does not match any other rules,
$\mathtt{DeclSub}$ can be applied and try to synthesize a type
before checking whether the synthesized one is more general.

Lambda-functions is a critical component of bidirectional type system, and they require much attention.
Checking $\lam e$ against function type $A\to B$ is simple:
we check the body $e$ against $B$ under the context extended with $x:A$.
However, synthesizing a lambda-function is not easy in general,
and this \emph{predicative} declarative system is able to
synthesize to a suitable monotype $\sigma\to\tau$.
In other words, if no type information is given for a lambda function,
we will synthesize it to the best mono-typed function according to the context,
which also requires some guessing similar to the subtyping rule $\mathtt{\forall L}$.

Application $(e_1~e_2)$ is handled by Rule $\mathtt{Decl{\to}E}$,
which firstly synthesize type $A$ for the function $e_1$.
If $A$ is a function type $B\to C$, Rule $\mathtt{Decl{\to}App}$ is applied,
by checking the argument $e_2$ against $B$ and returns type $C$.
The only other possibile kind of types that
the function $e_1$ could be synthesized is polymorphic types $\all A$.
In that case, we instantiate $A$ to $[\tau/a]A$ with a monotype $\tau$ (which is also guessed)
according to Rule $\mathtt{Decl{\to}I{\To}}$.
If $[\tau/a]A$ is a function type, Rule $\mathtt{Decl{\to}App}$ proceeds;
if $[\tau/a]A$ is still polymorphic, Rule $\mathtt{Decl{\to}I{\To}}$ is recursively applied.

@@

\paragraph{Overlapping Rules}
\jimmy{Mention the resolution of overlapping briefly, and then point to algorithm and metatheory}
Some of the declarative rules overlaps with each other.
Declarative subtyping rules $\mathtt{{\le}\forall L}$ and $\mathtt{{\le}\forall R}$
both match the conclusion $\Psi\vdash \all A \le \all B$.
In such case, choose $\mathtt{{\le}\forall R}$ first is always better,
since we introduce the type variable $a$ to the context earlier,
which gives more flexibility on the choice of $\tau$.

@@

The DK algorithm revolves around their notion of \emph{algorithmic context}.
\[
\begin{array}{l@{\qquad}lcl}
\text{Algorithmic Contexts}\qquad&\Gamma,\Delta,\Theta &::=&\quad \nil \mid
\Gamma, a \mid \Gamma, \hat{\alpha} \mid \Gamma, \hat{\alpha} = \tau \mid
\Gamma, \blacktriangleright_{\hat{\alpha}}
\end{array}
\]
In addition to the regular (universally quantified) type variables $a$, the
algorithmic context also contains \emph{existential} type variables
$\hat{\alpha}$. These are placeholders for monotypes $\tau$ that are still to
be determined by the inference algorithm. When the existential variable is
``solved'', its entry in the context is replaced by the assignment
$\hat{\alpha} = \tau$. With the notation $[\Gamma]$ an algorithmic context
$\Gamma$ denotes the substitution that replaces all its solved existential type
variables with their solution.

@@

The sequential order of entries in the algoritmic context, in combination with
the threading of contexts,  does not perfectly capture the scoping of all
existential variables. For this reason the DK algorithm uses scope markers
$\blacktriangleright_{\hat{\alpha}}$ in a few places. An example is given in the following
rule:
\[
\inferrule*
  {\Gamma,\blacktriangleright_{\hat{\alpha}}, \hat{\alpha} \vdash [\hat{\alpha}/a]A \le B \dashv \Delta,\blacktriangleright_{\hat{\alpha}},\Theta}
  {\Gamma \vdash \forall a. A \le B \dashv \Delta}
\]
To indicate that the scope of $\hat{\alpha}$ is local to the subtyping check
$[\hat{\alpha}/a]A \le B$, the marker is pushed onto its input stack and popped
from the output stack together with the subsequent part $\Theta$, which may
refer to $\hat{\alpha}$. (Remember that later entries may refer to earlier
ones, but not vice versa.) This way $\hat{\alpha}$ does not escape its scope.

@@

More seriously, we have also found a lemma that simply does not hold, 
Lemma 29 (Parallel Admissibility). See Appendix~\ref{} for a detailed explanation
and counterexample. This lemma is a cornerstone of the two metatheoretical results 
of the algorithm, soundness and completeness with respect to the declarative system.
In particular, both instantiation soundness (i.e. a part of subtyping
soundness) and typing completeness both directly require the broken lemma.
Moreover, Lemma 54 (Typing Extension) also requires the broken lemma and is
itself used 13 times in the proof of typing soundness and completeness.
Unfortunately, we have not yet found a way to fix this problem.
\tom{You have communicated this to the authors and not received a fix, right?
     We could mention that too then.}

@@

Besides the problematic lemmas in the DK metatheory, there are other reasons to
look for an alternative algorithmic formulation of the type system that is more
amenable to mechanisation. Indeed, two aspects that are particularly
challenging to mechanise are the scoping of universal and existential type
variables, and the propagation of the instantiation of existential type
variables across judgments.

>>>>

This section starts by introducing DK's declarative type system. Then
it discusses several techniques that have been used in algorithmic
formulations, which have influenced our own algorithmic design.
Finally we introduce the novelties of our new algorithm.
In particular the support for inference judgements in
judgement worklists, and a new form of judgement worklist
that the unifies \emph{ordered contexts} and the worklists themselves.

@@

\paragraph{Syntax.}
The syntax of DK's declarative system~\cite{dunfield2013complete} is shown in Figure~\ref{fig:decl:syntax}.
A declarative type $A$ is either the unit type $1$, a type variable $a$,
a universal quantification $\forall a. A$ or a function type $A \to B$.
Nested universal quantifiers are allowed for types,
but monotypes $\tau$ do not have any universal quantifier.
Terms include a unit term $()$, variables $x$, lambda-functions $\lam e$,
applications $e_1~e_2$ and annotations $(e:A)$.
Contexts $\Psi$ are sequences of type variable declarations and
term variables with its type declared $x:A$.

@@

\begin{figure}[t]
\begin{tabular}{rl}
    \framebox{$\Psi \vdash A \Lto B$} & $e$ checks against input type $A$.\\[0.5mm]
    \framebox{$\Psi \vdash A \To B$} & $e$ synthesizes output type $A$.\\[0.5mm]
    \framebox{$\Psi \vdash \appInf{A}{e}{C}$} & Applying a function of type $A$ to $e$ synthesizes type $C$.
\end{tabular}
\begin{gather*}
\inferrule*[right=$\mathtt{DeclVar}$]
    {(x:A)\in\Psi}{\Psi\vdash x\To A}
\qquad
\inferrule*[right=$\mathtt{DeclSub}$]
%e \neq \lam e' \quad B \neq \all B' \quad 
    {\Psi\vdash e\To A \quad \Psi\vdash A\le B}
    {\Psi \vdash e\Lto B}
\\
\inferrule*[right=$\mathtt{DeclAnno}$]
    {\Psi \vdash A \quad \Psi\vdash e\Lto A}
    {\Psi\vdash (e:A)\To A}
\qquad
\inferrule*[right=$\mathtt{Decl1I}$]
    {~}{\Psi\vdash () \Lto 1}
\qquad
\inferrule*[right=$\mathtt{Decl1I\To}$]
    {~}{\Psi\vdash () \To 1}
\\
\inferrule*[right=$\mathtt{Decl\forall I}$]
    {\Psi,a \vdash e \Lto A}
    {\Psi\vdash e\Lto \all A}
\qquad
\inferrule*[right=$\mathtt{Decl\forall App}$]
    {\Psi \vdash \tau \quad \Psi\vdash \appInf{[\tau/a]A}{e}{C} }
    {\Psi\vdash \appInf{\all A}{e}{C}}
\\
\inferrule*[right=$\mathtt{Decl\to I}$]
    {\Psi,x:A \vdash e\Lto B}
    {\Psi\vdash \lam e \Lto A \to B}
\qquad
\inferrule*[right=$\mathtt{Decl\to I\To}$]
    {\Psi\vdash \sigma\to\tau \quad \Psi,x:\sigma \vdash e\Lto \tau}
    {\Psi\vdash \lam e \To \sigma\to\tau}
\\
\inferrule*[right=$\mathtt{Decl\to E}$]
    {\Psi\vdash e_1\To A \quad \Psi\vdash \appInf{A}{e_2}{C}}
    {\Psi\vdash e_1~e_2 \To C}
\qquad
\inferrule*[right=$\mathtt{Decl\to App}$]
    {\Psi\vdash e \Lto A}
    {\Psi\vdash \appInf{A \to C}{e}{C}}
\end{gather*}
\caption{Declarative Typing}\label{fig:decl:typing}
\end{figure}
\bruno{Some of the rule names are messed up.}

%The DK subtyping relation was adopted from \citet{odersky1996putting}.

@@

The most interesting rule is $\mathtt{{\le}\forall L}$.
If some instantiation of $\all A$, $[\tau/a]A$, is a subtype of $B$,
then $\all A \le B$.
The monotype $\tau$ we used to instantiate $a$ is \emph{guessed} in this
declarative rule, but the algorithmic system does not guess and defers the
instantiation until it can determine the monotype deterministically.
The fact that $\tau$ is a monotype rules out the possibility of
polymorphic (or impredicative) instantiation.
However this restriction ensures that the subtyping relation remains
decidable. Allowing an arbitrary type (rather than a monotype) in rule $\mathtt{{\le}\forall L}$
is know to give rise to an undecidable subtyping relation~\cite{tiuryn1996subtyping}.
Peyton Jones et al.~\cite{jones2007practical} also impose the restriction of
predicative instantiation in their type system.
Both systems are adopted by several practical programming languages.

@@

Some of the rules handle simple cases with common approaches.
Rule $\mathtt{DeclVar}$ looks term variables up in the context.
Rules $\mathtt{Decl1I}$ and $\mathtt{Decl1I{\To}}$ checks and synthesizes the unit type, respectively.
Rule $\mathtt{DeclAnno}$ synthesizes the type of the annotated expression $(e:A)$
to its annotated type $A$, and requires $e$ to be checked against $A$.
Checking an expression $e$ against a polymorphic type $\all A$ under the context $\Psi$ succeeds
if $e$ checks against A under an extended context $(\Psi, a)$, where $a$ is picked as a free variable.

The subsumption rule $\mathtt{DeclSub}$ depends on the subtyping relation,
and changes the mode from checking to synthesis: if $e$ synthesizes $A$ and $A\le B$
($A$ is more polymorphic than $B$), then $e$ can be checked against $B$.
If a checking problem does not match any other rules,
$\mathtt{DeclSub}$ can be applied and try to synthesize a type
before checking whether the synthesized one is more general.

Lambdas are critical components of the bidirectional type system, and they require much attention.
Checking $\lam e$ against function type $A\to B$ is simple:
we check the body $e$ against $B$ under the context extended with $x:A$.
However, synthesizing a the type of an unannotated lambda is not easy. 
Nevertheless the \emph{predicative} declarative system is able to
synthesize to a suitable monotype $\sigma\to\tau$, if one exists.
In other words, if no type information is given for a lambda function,
we will synthesize it to the best mono-typed function according to the context,
which also requires some guessing similar to the subtyping rule $\mathtt{\forall L}$.

Applications $e_1~e_2$ are handled by Rule $\mathtt{Decl{\to}E}$,
which firstly synthesizes type $A$ for the function $e_1$.
If $A$ is a function type $B\to C$, Rule $\mathtt{Decl{\to}App}$ is applied,
by checking the argument $e_2$ against $B$ and returns type $C$.
The only other possibile kind of type that
the function $e_1$ could be synthesized to is a polymorphic type $\all A$.
In that case, we instantiate $A$ to $[\tau/a]A$ with a monotype $\tau$ (which is also guessed)
according to Rule $\mathtt{Decl{\to}I{\To}}$.
If $[\tau/a]A$ is a function type, Rule $\mathtt{Decl{\to}App}$ proceeds;
if $[\tau/a]A$ is still polymorphic, Rule $\mathtt{Decl{\to}I{\To}}$ is recursively applied.

@@

\paragraph{Overlapping Rules}
\jimmy{Mention the resolution of overlapping briefly, and then point to algorithm and metatheory}
Some of the declarative rules overlaps with each other.
Declarative subtyping rules $\mathtt{{\le}\forall L}$ and $\mathtt{{\le}\forall R}$
both match the conclusion $\Psi\vdash \all A \le \all B$.
In such case, choosing $\mathtt{{\le}\forall R}$ first is always better,
since we introduce the type variable $a$ to the context earlier,
which gives more flexibility on the choice of $\tau$.

@@

DK's algorithm version revolves around their notion of \emph{algorithmic context}.
\[
\begin{array}{l@{\qquad}lcl}
\text{Algorithmic Contexts}\qquad&\Gamma,\Delta,\Theta &::=&\quad \nil \mid
\Gamma, a \mid \Gamma, \hat{\alpha} \mid \Gamma, \hat{\alpha} = \tau \mid
\Gamma, \blacktriangleright_{\hat{\alpha}}
\end{array}
\]
In addition to the regular (universally quantified) type variables $a$, the
algorithmic context also contains \emph{existential} type variables
$\hat{\alpha}$. These are placeholders for monotypes $\tau$ that are still to
be determined by the inference algorithm. When the existential variable is
``solved'', its entry in the context is replaced by the assignment
$\hat{\alpha} = \tau$. With the notation $[\Gamma]$ an algorithmic context
$\Gamma$ denotes the substitution that replaces all its solved existential type
variables with their solution.

@@

\paragraph{Markers for scoping.}
The sequential order of entries in the algoritmic context, in combination with
the threading of contexts,  does not perfectly capture the scoping of all
existential variables. For this reason the DK algorithm uses scope markers
$\blacktriangleright_{\hat{\alpha}}$ in a few places. An example is given in the following
rule:
\[
\inferrule*
  {\Gamma,\blacktriangleright_{\hat{\alpha}}, \hat{\alpha} \vdash [\hat{\alpha}/a]A \le B \dashv \Delta,\blacktriangleright_{\hat{\alpha}},\Theta}
  {\Gamma \vdash \forall a. A \le B \dashv \Delta}
\]
To indicate that the scope of $\hat{\alpha}$ is local to the subtyping check
$[\hat{\alpha}/a]A \le B$, the marker is pushed onto its input stack and popped
from the output stack together with the subsequent part $\Theta$, which may
refer to $\hat{\alpha}$. (Remember that later entries may refer to earlier
ones, but not vice versa.) This way $\hat{\alpha}$ does not escape its scope.

@@

More seriously, we have also found a lemma that simply does not hold, 
Lemma 29 (Parallel Admissibility). See Appendix~\ref{} for a detailed explanation
and counterexample. This lemma is a cornerstone of the two metatheoretical results 
of the algorithm, soundness and completeness with respect to the declarative system.
In particular, both instantiation soundness (i.e. a part of subtyping
soundness) and typing completeness directly require the broken lemma.
Moreover, Lemma 54 (Typing Extension) also requires the broken lemma and is
itself used 13 times in the proof of typing soundness and completeness.
Unfortunately, we have not yet found a way to fix this problem.
\footnote{We found the issue with Lemma 29 in 2016 in an earlier attempt to mechanically
  formalize DK's algorithm. The authors acknowledged the problem after we contacted them through email.
  Although they briefly mentioned that it should be possible to use a weaker lemma instead they did
  not go into details.
}
\bruno{We should acknowledge Ningning for finding this problem, but I think
  this may be against the anonymous submission? Perhaps we do not mention her name
in the submission, and if accepted we mention her?}
\tom{You have communicated this to the authors and not received a fix, right?
     We could mention that too then.}

@@

Besides the problematic lemmas in DK's metatheory, there are other reasons to
look for an alternative algorithmic formulation of the type system that is more
amenable to mechanisation. Indeed, two aspects that are particularly
challenging to mechanise are the scoping of universal and existential type
variables, and the propagation of the instantiation of existential type
variables across judgments.
